Tactic Script Optimisation for Aesop

### 2024-09-14 paper.tex
Clarify that naive example script happens to be depth-first
Replace wrong ext example with split example

Unfortunately, when we extract the tactics from an Aesop proof tree, they are generally not in depth-first order, for two reasons.
First, a rule can generate multiple tactics and it would be onerous to require that these are always in depth-first order.
Second, Aesop runs the rules in best-first order rather than depth-first, and the proof tree reflects this.

@@

As a less straightforward example, the \tac{ext} tactic also requires special treatment.
This tactic applies extensionality rules, which reduce an equality of two objects to equalities of different objects (often subcomponents).
During this process, the tactic can introduce new hypotheses, but it provides no way to name these, which would be required to preserve the invariant that there are no unnamed hypotheses.
To address this issue, Aesop implements a variant of \tac{ext} that applies extensionality lemmas as long as the application of each lemma generates only one subgoal.
At this point, it generates a step with tactic $\tac{ext} : n$.
This limits \tac{ext} to depth $n$, with $n$ being the number of lemmas applied so far, guaranteeing that the tactic applies precisely the lemmas applied by the rule.
Then Aesop adds a \tac{rename\_i} step that gives names to the hypotheses introduced so far and repeats the process until no extensionality lemmas are applicable any more.
Again, most of these \tac{rename\_i} calls will be removed in Section~\ref{sec:global-optimisation}.

@@

\paragraph{Focused rename\_i}
The first pass applies the following rewrite rule from left to right wherever the left-hand side occurs in the script.
\[
  \arraycolsep=0pt
  \begin{array}{ll}
    \tacbullet &\tac{rename\_i}~x₁ \dots xₙ \\
               &s
  \end{array}
  \quad\Longrightarrow\quad
  \begin{array}{ll}
    &\tac{next}~x₁ \dots xₙ ⇒ \\
    &\quad s
  \end{array}
\]
Both the left-hand and the right-hand tactic expressions focus on the first goal, then rename the last $n$ inaccessible hypotheses in the local context and continue with the tactic sequence $s$.
The right-hand expression uses the $\tac{next}$ combinator, which is designed specifically for this situation and is thus considered more idiomatic.
But it cannot be generated directly: when a tactic generates its script steps, it does not know whether its subgoals are focusable, and hence whether $\tac{next}$ is applicable.
The tactic therefore generates a $\tac{rename\_i}$ script step and if the goal later turns out to be focusable, a bullet is added during structuring.
The static optimisation pass then cleans up the script, replacing bullet and $\tac{rename\_i}$ with $\tac{next}$.
This is a considerable improvement since both the $\tac{ext}$ rule, discussed in Section~\ref{sec:generation}, and another rule that runs the \tac{split} tactic produce many focusable \tac{rename\_i} calls.

>>>>

Unfortunately, when we extract the tactics from an Aesop proof tree, they are generally not in depth-first order, for two reasons.
First, a rule can generate multiple tactics and it would be onerous to require that these are always in depth-first order.
Second, Aesop runs the rules in best-first order rather than depth-first, and the proof tree reflects this.
(The script from Figure~\ref{fig:big-example-unoptimised} happens to be in depth-first order already.)

@@

Another example concerns an Aesop rule that runs the \tac{split} tactic.
This tactic performs case analysis on the discriminants of if-then-else and match expressions in the goal.
Choices made during the case split are reflected in new hypotheses, but \tac{split} provides no mechanism to name these hypotheses, so they are inaccessible.
Since inaccessible hypotheses are forbidden, we must add a call to \tac{rename\_i} to each \tac{split} step.
However, these calls are usually replaced with more idiomatic constructs by the post-processing step described in Section~\ref{sec:global-optimisation}.

@@

\paragraph{Focused rename\_i}
The first pass applies the following rewrite rule from left to right wherever the left-hand side occurs in the script.
\[
  \arraycolsep=0pt
  \begin{array}{ll}
    \tacbullet &\tac{rename\_i}~x₁ \dots xₙ \\
               &s
  \end{array}
  \quad\Longrightarrow\quad
  \begin{array}{ll}
    &\tac{next}~x₁ \dots xₙ ⇒ \\
    &\quad s
  \end{array}
\]
Both the left-hand and the right-hand tactic expressions focus on the first goal, then rename the last $n$ inaccessible hypotheses in the local context and continue with the tactic sequence $s$.
The right-hand expression uses the $\tac{next}$ combinator, which is designed specifically for this situation and is thus considered more idiomatic.
But it cannot be generated directly: when a tactic generates its script steps, it does not know whether its subgoals are focusable, and hence whether $\tac{next}$ is applicable.
The tactic therefore generates a $\tac{rename\_i}$ script step and if the goal later turns out to be focusable, a bullet is added during structuring.
The static optimisation pass then cleans up the script, replacing bullet and $\tac{rename\_i}$ with $\tac{next}$.
This is a considerable improvement since the \tac{split} rule discussed in Section~\ref{sec:generation} produces many focusable \tac{rename\_i} calls.
